Potential of I/O-Aware Workflows in Climate and Weather

### 2020-02-20 paper.tex
Minor changes (Dave) and review of related work / conclusion

\usepackage{array}
\usepackage{todonotes}
\usepackage[capitalize]{cleveref}
\Crefname{lstlisting}{Listing}{Listings}
\usepackage{subcaption}
\usepackage{comment}
\usepackage{float}
\usepackage{listings}
\usepackage{textcomp} %for upquote=true (straight quotes)
\definecolor{pblue}{rgb}{0.13,0.13,1}
\definecolor{pgreen}{rgb}{0,0.5,0}
\definecolor{pred}{rgb}{0.9,0,0}
\definecolor{pgrey}{rgb}{0.46,0.45,0.48}
\lstset{
    basicstyle=\ttfamily\footnotesize,
    numberstyle=\ttfamily\tiny,
    frame=single,
    numbers=left,
    numbersep=1em,
    xleftmargin=2em,
    language=XML, % to avoid type being blue
    breaklines=true,
    breakatwhitespace=true,
    postbreak=\hbox{$\hookrightarrow$ },
    showstringspaces=false,
    tabsize=2,
    commentstyle=\color{pgreen},
    keywordstyle=\color{pblue},
    stringstyle=\color{pred},
    moredelim=[il][\textcolor{pgrey}]{\[\]},
    moredelim=[is][\textcolor{pgrey}]{\%\%}{\%\%},
  columns=fullflexible,
    upquote=true
}

@@

  \item \textbf{Script} executes the shell commands which is one or multiple (potentially parallel) applications to run sequentially.
  During the creation of the script, Cylc has included variables that describe the task in the workflow.
  The user script may use commands to create a filename considering the cycle of Cylc.
  The information is typically fed into the application(s) representing the task, and so defines the storage location.

@@

\textbf{Technology.}
Manual tiering requires the user or application to control the data placement, i.e., storing data (typically in the form of files on a particular storage system) and, usually, moving data between storage by scripts.
One limitation of such an approach is that decisions about how data are mapped and packaged into files are made by the producing application, and cannot be changed without manual intervention by a downstream application.

@@

\textbf{Workflows. }
A good overview of the flavours of Scientific Workflow Management Systems (SWfMS) and their application to data-intensive workflows is given in \cite{liu2015survey}.
The article states a SWfMS "should enable the parallel execution of data-intensive scientific workflows and exploit large amounts of distributed resources".
%"An abstract scientific workflow is defined by the functionality of each activity (or sub-workflow) and data dependencies between activities (or sub-workflows)"
Existing solutions recognize challenges in data variety (formats of the input data), the opportunity to optimize the schedule by moving code to data, and to consider capacity of available data storage.
There have been also solutions that allow to specify the data dependencies for tasks.
The execution engine Dryad~\cite{isard2007dryad}, for example, allows to transfer data between tasks via files or directly using TCP connections and attempts to schedule tasks on the same nodes or racks.
Swift/T is a scripting language for describing dataflow processing allowing to execute ensembles of applications~\cite{ozik2016desktop}.
Recent improvements aim to migrate data to a local cache allowing to exploit locality, e.g., in \cite{dai2018cross}, information about locality is proposed to be stored in extended attributes.
In \cite{TUIBIHWLSC19}, an approach was presented to monitor and analyze IO behavior of HPC workflows.

Various early research covers grid workflows and lately cloud use cases and attempts to maximize data locality in that respect.
Economic factors (including storage costs) for workflow execution are discussed in \cite{alkhanak2016cost}.
In \cite{deelman2019role}, the authors discuss the role of Machine Learning for workflow execution and elaborate an the general potential for resource provisioning such as optimization of runtime parameters, data movements, and hierarchical storage.
An ML model is introduced in \cite{subedi2019leveraging} that stages data for in-situ analysis by exploiting the access patterns.

@@

Organising the data placement on storage tiers is typically performed manually (and particularly in the domain of climate and weather) by the users or via policies, leading to suboptimal optimisations.
However, the manual optimization and hard-coding of storage locations is non-portable and an error-prone task.
We believe users must be able to express their workflows in an abstract fashion.
By increasing the abstraction level for scientists, not only tedious manual optimisation could be automatised, but also optimization strategies for data placement and data reduction can be harnessed.
With knowledge about the data pattern, the runtime system could generate (near-)optimal execution plans and monitor their execution.
In this work, we describe the general vision and a specific design for the software stack in the domain in climate and weather that we work on in the ESiWACE project.
The changes proposed increase the opportunity for smarter scheduling of storage in heterogeneous storage environments by considering the characteristics of data and system in a workflow.

@@

\small
This project is funded by the European Union's Horizon 2020 research and innovation programme under grant agreement No. 823988.

@@

>>>>

\usepackage{array}
\usepackage{todonotes}
\usepackage[capitalize]{cleveref}
\Crefname{lstlisting}{Listing}{Listings}
\usepackage{subcaption}
\usepackage{comment}
\usepackage{float}
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{textcomp} %for upquote=true (straight quotes)
\definecolor{pblue}{rgb}{0.13,0.13,1}
\definecolor{pgreen}{rgb}{0,0.5,0}
\definecolor{pred}{rgb}{0.9,0,0}
\definecolor{pgrey}{rgb}{0.46,0.45,0.48}
\lstset{
    basicstyle=\ttfamily\footnotesize,
    numberstyle=\ttfamily\tiny,
    frame=single,
    numbers=left,
    numbersep=1em,
    xleftmargin=2em,
    language=XML, % to avoid type being blue
    breaklines=true,
    breakatwhitespace=true,
    postbreak=\hbox{$\hookrightarrow$ },
    showstringspaces=false,
    tabsize=2,
    commentstyle=\color{pgreen},
    keywordstyle=\color{pblue},
    stringstyle=\color{pred},
    moredelim=[il][\textcolor{pgrey}]{\[\]},
    moredelim=[is][\textcolor{pgrey}]{\%\%}{\%\%},
  columns=fullflexible,
    upquote=true
}

@@

  \item \textbf{Script} executes the shell commands which is one or multiple (potentially parallel) applications to run sequentially.
  During the creation of the script, Cylc has included variables that describe the task in the workflow.
  The user script may use commands to create a filename considering the cycle of Cylc or store data in a workflow-specific shared directory.
  The information is typically fed into the application(s) representing the task, and so defines the storage location.

@@

\textbf{Technology.}
Manual tiering requires the user or application to control the data placement, i.e., storing data typically in the form of files on a particular storage system and, usually, moving data between storage by scripts.
One limitation of such an approach is that decisions about how data are mapped and packaged into files are made by the producing application, and cannot be changed without manual intervention by a downstream application.

@@

\textbf{Workflows. }
A good overview of the flavours of Scientific Workflow Management Systems (SWfMS) and their application to data-intensive workflows is given in \cite{liu2015survey}.
The article states a SWfMS \textit{“should enable the parallel execution of data-intensive scientific workflows and exploit large amounts of distributed resources”}.
%"An abstract scientific workflow is defined by the functionality of each activity (or sub-workflow) and data dependencies between activities (or sub-workflows)"
Existing solutions recognize challenges in data variety (formats of the input data), the opportunity to optimize the schedule by moving code to data, and may consider capacity of available data storage.
There have been solutions that allow to specify the data dependencies for tasks.
The execution engine Dryad~\cite{isard2007dryad}, for example, allows to transfer data between tasks via files or directly using TCP connections and attempts to schedule tasks on the same nodes or racks.
Swift/T is a scripting language for describing dataflow processing allowing to execute ensembles of applications~\cite{ozik2016desktop}.
Recent improvements aim to migrate data to a local cache allowing to exploit locality, e.g., in \cite{dai2018cross}, information about locality is proposed to be stored in extended attributes.
In \cite{TUIBIHWLSC19}, an approach was presented to monitor and analyze IO behavior of HPC workflows.

Various early research in grid workflows and lately cloud use cases attempts to maximize data locality in that respect.
Economic factors (including storage costs) for workflow execution are discussed in \cite{alkhanak2016cost}.
In \cite{deelman2019role}, the authors discuss the role of Machine Learning for workflow execution and elaborate an the general potential for resource provisioning such as optimization of runtime parameters, data movements, and hierarchical storage.
An ML model is introduced in \cite{subedi2019leveraging} that stages data for in-situ analysis by exploiting the access patterns.

@@

Organising the data placement on storage tiers is particularly in the domain of climate and weather performed by the users or via policies, leading often to suboptimal decisions.
Additionally, the manual optimization and hard-coding of storage locations is non-portable and an error-prone task.
We believe users must be able to express their workflows in an abstract fashion.
By increasing the abstraction level for scientists, not only tedious manual optimisation could be automatised, but also strategies for data placement and data reduction can be harnessed.
With knowledge about the data pattern, the runtime system could generate optimized execution plans and monitor their execution.
In this work, we describe the general vision and a specific design for the software stack in the domain in climate and weather that we work on in the ESiWACE project.
The changes proposed increase the opportunity for smarter scheduling of storage in heterogeneous storage environments by considering the characteristics of data and system in a workflow.

@@

\small
This project is funded by the European Union's Horizon 2020 research and innovation programme under grant agreement No. 823988.
We thank our collaborator Bryan Lawrence for their input to this article.

@@

CYLC_SUITE_SHARE_DIR
Typically the data generated by cylc workflows is stored in the “share” directory which is referenced by an environment variable rather than a command:
https://cylc.github.io/doc/built-sphinx/suite-config.html#suite-share-directories
